---
title: "North Pacific Pelagic Seabird Database"
author: "Alberto Rovellini"
date: "9/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(maps)
library(mapdata)
library(rbgm)
library(viridis)
```

```{r}
select <- dplyr::select
```


This document reads and explores the North Pacific Pelagic Seabird Database, and maps it to the Atlantis GOA geometry. Worth noting that this data includes different sampling techniques, which we may need to consider if we do any SDM-ing or biomass standardization in space.

# Read data

Read the data, the transect information, and the Atlantis BGM.
```{r}
dat <- read.csv('../data/NPPSD_v3.0/tbl_DATA_OBS.csv')
tows <- read.csv('../data/NPPSD_v3.0/tbl_LOCATION.csv')
# and BGM
atlantis_bgm <- read_bgm('../data/GOA_WGS84_V4_final.bgm')
atlantis_box <- atlantis_bgm %>% box_sf()
# utilities
atlantis_crs <- atlantis_bgm$extra$projection
atlantis_bbox <- atlantis_box %>% st_bbox()
```

Keep only Laysan and black-footed albatross, and Sooty shearwaters.
```{r}
nonbreeding <- c('LAAL','BFAL','SOSH')
dat_nb <- dat %>% filter(NPPSD.4.Letter.Code %in% nonbreeding)
```

```{r}
dat_join <- dat_nb %>% select(Master.Key,Common.Name,NPPSD.4.Letter.Code,Density) %>% 
  left_join(tows %>% select(Master.Key,Lon,Lat,Year), by = 'Master.Key') %>%
  mutate(Density_num=as.numeric(Density)) %>%
  drop_na() # remove NAs 
```

# Map to Atlantis groups

We need to match species with those that we have from Guenette (2005). Let's see how much gets lost if we do that, in terms of abundance in the GOA. Hopefully not too much. Check that there are not nomenclature mismatches between species, and fix them if there are.
```{r}
# guenette <- read.csv('../data/Guenette.csv', fileEncoding = 'UTF-8-BOM')
# guenette <- guenette %>% select(Common.name,Scientific.name,Proposed.group) %>% mutate(Common.name=toupper(Common.name))
# guenette <- guenette[-nrow(guenette),]
# guenette <- guenette %>% mutate(Proposed.group=str_replace(Proposed.group,', ', '_'))
# 
# dat_guenette <- dat_join %>% mutate(Common.Name=toupper(Common.Name)) %>% 
#   left_join(guenette,by=c('Common.Name'='Common.name')) %>%
#   drop_na()

# now group by event and by proposed group and sum densities
dat_groups <- dat_join %>% st_as_sf(coords = c('Lon','Lat'), crs = 4326) %>%
  st_transform(crs = atlantis_crs)
```

Draw some maps. Beware that density is not available for all transects because some transects do not have area estimates.
```{r}
# get some coastlines
coast <- maps::map(database = 'worldHires', regions = c('USA','Canada'), plot = FALSE, fill=TRUE)
coast_sf <- coast %>% st_as_sf(crs = 4326) %>% st_transform(crs = atlantis_crs) %>% st_combine()

dat_groups %>% 
  st_crop(atlantis_bbox) %>% # limit this to the plotting region
  #filter(Year==2017) %>%
  ggplot()+
  geom_sf(aes(color=log1p(Density_num)))+
  scale_color_viridis()+
  geom_sf(data=coast_sf)+
  coord_sf(xlim=c(atlantis_bbox$xmin, atlantis_bbox$xmax),ylim=c(atlantis_bbox$ymin, atlantis_bbox$ymax+50000))+
  theme_bw()+
  facet_wrap(~Common.Name, ncol=2)+
  labs(title='Bird density from the North Pacific Pelagic Seabird Database')
```

It is a pretty sparse dataset, compared to the bottom trawl data. It does build up if one ignores the year factor though - so much so that plotting becomes an issue. 

## Overlap with Atlantis

Let's subset to the data points that overlap with our model domain.
```{r}
dat_groups_atlantis <- dat_groups %>% 
  mutate(Lon = st_coordinates(dat_groups)[,1],
         Lat = st_coordinates(dat_groups)[,2]) %>%
  st_join(atlantis_box %>% select(.bx0)) %>% 
  drop_na() %>%
  select(-geometry) %>%
  st_as_sf(coords=c('Lon','Lat'), crs = atlantis_crs)

dat_groups_atlantis %>%
  ggplot()+
  geom_sf(aes(color=log1p(Density_num)))+
  scale_color_viridis()+
  geom_sf(data=coast_sf)+
  coord_sf(xlim=c(atlantis_bbox$xmin, atlantis_bbox$xmax),ylim=c(atlantis_bbox$ymin, atlantis_bbox$ymax))+
  theme_bw()+
  facet_wrap(~Common.Name, ncol=2)+
  labs(title='Bird density from the North Pacific Pelagic Seabird Database')
```

# Prepare data for sdmTMB

Need to do a few things here:

1. Turn coordinates from m to km for numerical stability in sdmTMB.
2. Calculate distance of each point from shore.
```{r}
dat_groups_atlantis <- dat_groups %>% 
  mutate(lon = st_coordinates(dat_groups)[,1]/1000, # prepare coordinates in km to aid numerical stability in sdmTMB
         lat = st_coordinates(dat_groups)[,2]/1000) %>%
  st_join(atlantis_box %>% select(.bx0)) %>% 
  drop_na() %>%
  mutate(distance = as.vector(st_distance(geometry,coast_sf))/1000) %>%
  st_set_geometry(NULL) 

# do the same for the grid that sdmTMB will predict on
load("../data/atlantis_grid_depth.Rdata")

atlantis_grid_dist <- atlantis_grid_depth %>% 
  select(-depth,-insideY,-insideX) %>% 
  st_as_sf(coords=c("x","y"), crs = atlantis_crs) %>%
  mutate(distance = as.vector(st_distance(geometry,coast_sf))/1000) %>%
  mutate(lon=st_coordinates(geometry)[,1]/1000,lat=st_coordinates(geometry)[,2]/1000) %>%
  st_set_geometry(NULL)
```

Visualise.
```{r}
dat_groups_atlantis %>% ggplot()+
  geom_point(aes(x=lon,y=lat,color=distance))+
  scale_color_viridis()+
  theme_minimal()+
  labs(title = 'Distance from shore of observations in the NPPSD')


atlantis_grid_dist %>% ggplot()+
  geom_point(aes(x=lon,y=lat,color=distance))+
  scale_color_viridis()+
  theme_minimal()+
  labs(title = 'Distance from shore of points in the regular prediction grid')
```
Seems sensible.

3. What about zero counts in some hauls? For each haul, we need counts of all four groups, i.e. we need to pad all hauls with zeroes to use SDMs. Do this one species at a time as it is much easier and otherwise.
```{r}
# make a frame with all transects and empty density and species info
all_obs <- dat %>% select(Master.Key,Common.Name,NPPSD.4.Letter.Code,Density) %>% 
  left_join(tows %>% select(Master.Key,Lon,Lat,Year), by = 'Master.Key') %>%
  mutate(Density_num=as.numeric(Density)) %>%
  drop_na() %>%
  mutate(Common.Name=NA, NPPSD.4.Letter.Code=NA, Density=0, Density_num=0) %>% # set groups to NA and densities to 0
  distinct() %>%
  st_as_sf(coords = c('Lon','Lat'), crs = 4326) %>% 
  st_transform(crs = atlantis_crs) 
all_obs <- all_obs %>% 
  mutate(lon = st_coordinates(all_obs)[,1]/1000, # prepare coordinates in km to aid numerical stability in sdmTMB
         lat = st_coordinates(all_obs)[,2]/1000) %>%
  st_join(atlantis_box %>% select(.bx0)) %>% 
  filter(!is.na(.bx0)) %>%
  mutate(distance = as.vector(st_distance(geometry,coast_sf))/1000) %>%
  st_set_geometry(NULL) 

# make a function that pulls one group out of the data and stitches transects with 0 observations
make_bird_data <- function(group) {
  this_group <- dat_groups_atlantis %>% filter(NPPSD.4.Letter.Code==group)
  # empty transects with respect to this group
  no_obs <- all_obs %>% filter(Master.Key %in% setdiff(all_obs$Master.Key,this_group$Master.Key))
  # bind this to observations
  this_group_complete <- rbind(this_group,no_obs) # row number checks out
  # fill the NAs in the Proposed.group column to match which group we are talking about
  this_group_complete <- this_group_complete %>% 
    select(-NPPSD.4.Letter.Code,-Density) %>%
    set_names(c('haul','name','year','dens','lon','lat','.bx0','distance'))
  # write this out
  write.csv(this_group_complete, paste0('../outputs/',group,'.csv'),row.names = FALSE)
}

purrr::map(nonbreeding,make_bird_data) 

```

Now move this to sdmTMB. This may work or not. Use as template the GOAIERP sdmTMB workflow as most similar.
